### 2014-10-23 Paper_v1.tex
Final draft version 1

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling blood coagulation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
In this study, we present a novel hybrid modeling approach which combines ordinary differential equation (ODE) modeling with logical rules to simulate an archetype biochemical network, 
the human coagulation cascade.  
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
This formulation was more than an order of magnitude smaller than current coagulation models, 
because many of the mechanistic details of coagulation were encoded as logical rules.
We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
We then validated the model ensemble using thrombin data sets that were not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems.

@@

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics, and encode mechanism.
However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, it is not clear whether we actually need precise estimates for all model parameters. 
Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well defined. 
Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
without requiring mechanism. \citep{Morris:2010aa}. 
Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
However, modeling complex dynamics with logical models is challenging. 
Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
while missing or incomplete mechanistic knowledge can be approximated using a logical approach.

In this study, we developed a hybrid approach which combines ODE modeling with logical rules to model a well studied biochemical network, 
the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by TF/FVIIa \cite{Butenas:2002aa}. 
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
leading to the formation of the tenase and prothrombinase complexes on activated platelets.
These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
neutralizes several of the proteases generated during coagulation, including thrombin.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

Coagulation models in the literature are typically formulated as systems of nonlinear ordinary differential equations, 
using mass-action or other kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
However, most of these previous studies were based upon detailed mechanistic knowledge. 
Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which mechanistic information is difficult or impossible to obtain. 
We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

%However, components could also correspond to non-modeled influences, categorial variables or other abstract quantities.

@@

The reduced order coagulation model captured the role of initial prothrombin abundance, and the decay of the thrombin signal following from ATIII activity (Fig. \ref{fig-diff_fII_NO_APC}).
However, we systematically under-predicted the thrombin peak and the strength of ATIII inhibition in this training data set.
On the other hand, with fixed thrombin parameters, we captured peak thrombin values and the decay of the thrombin signal (at least for the 150\% fII case) in the presence
of both ATIII and the protein C pathway (Fig. \ref{fig-diff_fII_APC}). Lastly, we were unable to capture global differences in initiation time \textit{across} separate data 
sets with a single ensemble of model parameters. These differences likely result from normal experimental variability.
For example different thrombin generation experiments within our 
training data (at the same physiological factor levels) have significantly different initiation times (supplemental results). 
However, this also highlights a potential shortcoming of the initiation module within the model. 
To capture the variability in initiation time \textit{across} training data sets, we included a constant time-delay parameter ($T_{D}$) for each data group.
The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. Introduction of the delay parameter allowed the model
to simulate multiple training data sets using a single ensemble of model parameters.
Taken together, the model identification results suggested 
that our kinetic-rules based approach could reproduce a panel of thrombin generation data sets conducted at physiological factor and inhibitors concentrations.
However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.

\subsection*{Validation of the reduced order coagulation model.}
We tested the predictive power of the reduced order coagulation model with validation data sets not used during model training. 
Two validation data sets were used, thrombin generation for various prothrombin and ATIII concentrations with the protein C pathway, 
and thrombin generation in normal versus hemophilic plasma in the presence of the protein C pathway. 
Lastly, we compared the qualitative output of the model to rFVIIa addition in the presence of hemophilia.
The hemophilia case was an especially difficult test as it was taken from a different study which used a plasma-based
\textit{in~vitro} assay involving platelets instead of phospholipid vesicles (PCPS). 
All kinetic and control parameters were fixed for the validation simulations. 
The only globally adjustable parameter $T_{D}$, was fixed within each validation data set but allowed to vary between data sets.
The reduced order model predicted the thrombin generation profile for ratios of prothrombin and ATIII in the absence of the protein C pathway (Fig. \ref{fig-fII_ATIII}).
Simulations near the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values (Fig. \ref{fig-fII_ATIII}B and C).
On the other hand, predictions for factors levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%), while qualitatively consistent with measured thrombin values, did show significant 
deviation from the measurements (Fig. \ref{fig-fII_ATIII}A and D).
Likewise, simulations of thrombin generation in normal versus hemophilia (missing both fVIII and fIX) were consistent with measured thrombin values (Fig. \ref{fig-hemophilia}).
We modeled the dependence of thrombin amplification on factor levels using a product rule ($\mathcal{Z} = fV\times fX\times fVIII\times fIX$), 
which was then was integrated using a $\min$ integration rule into the control variable governing amplification. 
Thus, in the absence of fVIII or fIX, the amplification control variable evaluated to zero, and the only thrombin produced was from initiation (Fig. \ref{fig-hemophilia}B).
However, the decay of the thrombin signal was underpredicted in the normal case (Fig. \ref{fig-hemophilia}A), while the activated thrombin level was overpredicted 
in hemophilia simulations, although thrombin generation was far less than normal (Fig. \ref{fig-hemophilia}B). 
Taken together, the reduced order model performed well in the physiological range of factors, even with
unmodeled components such as platelet activation in the hemophilia data set. Thus, our kinetic-rules framework predicted 
the output of a physiologically important cascades such as coagulation, despite significant unmodeled components.

The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa and rFVIIa, as we have only a single lumped initiation source (trigger).
Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin being produced was from initiation (Fig. \ref{fig-rFVII}, 1$\times$~trigger).
As we increased the strength of the trigger event, the thrombin peak time and the maximum value of thrombin increased (Fig. \ref{fig-rFVII}, 50$\times$~trigger). 
However, as the trigger strength increased, the thrombin generated quickly decayed.
For example, for the highest trigger strength simulated (50$\times$trigger), 95\% of the thrombin was gone by 20 min after initiation. 
We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig. \ref{fig-flux-analysis}).
The reduced order network includes four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
For normal coagulation, amplification and thrombin inhibition are the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
After 10 min, the dominate reaction has shifted to thrombin inhibition (both ATIII and APC action). In the current proof-of-principle model, 
APC inhibits upstream of amplification thus APC activity will slow the rate of thrombin formation directly from the initiation trigger. 
On the other hand, ATIII inhibits thrombin directly, as well as upstream coagulation factors.
In hemophilia (missing both fVIII and fIX), the amplification reaction does not occur and the only thrombin produced is from initiation (Fig. \ref{fig-flux-analysis}, center).
Initiation is quickly inhibited by APC, and the thrombin level stabilizes and eventually decays because of ATIII activity. 
Lastly, when 50$\times$rFVIIa is used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
initiation mechanisms dominate for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
However, similar to hemophilia alone, without amplification the thrombin signal is quickly extinguished by the combined action of ATIII and APC generated by thrombin.

@@

\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
We calculated the sensitivity of the time to maximum thrombin (peak time) and the thrombin exposure (area under the thrombin curve) 
for different levels of prothrombin, and protein C (Fig. \ref{fig-sensitivity}). 
Globally, 41\% of the parameters shifted in importance between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
The majority of these shifts were involved increased prothrombin and the protein C pathway, while only
5\% were directly associated with increased prothrombin alone. The rate constant for thrombin amplification was the most important parameter controlling the
peak thrombin time. While this parameter was differentially important for different prothrombin levels, and in the presence or absence of 
the activated protein C pathway, it's sensitivity was consistently above all other parameters in the model. The saturation constant governing thrombin amplification was
the second most important parameter, followed by the initiation control gain parameter. Other important parameters influencing the thrombin peak time included the control gain for
activated protein C formation, as well as the rate constant controlling ATIII inhibition of thrombin activity. 
On the other hand, only 27\% of the model parameters were differentially sensitive between the (\%fII,\%PC) (50\%, 0\%) and (150\%,100\%) cases 
for the thrombin exposure (Fig. \ref{fig-sensitivity}B). Of these parameters, all of the shifts were associated the interplay between thrombin formation and the protein C pathway. 
The rate constant controlling ATIII inhibition was the most important parameter controlling the thrombin exposure.
While this parameter was less important in the presence of protein C for 150\% prothrombin levels, it was significantly above all other parameters. 
Similar to the peak time, for 150\% prothrombin, the control gain for activated protein C formation was differentially important along with the rate constant controlling
amplification. However, the influence of the amplification parameter was much less in the thrombin exposure case versus peak time.

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Reduced order modeling of blood coagulation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
In this study, we present a novel hybrid modeling approach which combines ordinary differential equation (ODE) modeling with logical rules to simulate an archetype biochemical network, 
the human coagulation cascade.  
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
This formulation was more than an order of magnitude smaller than current coagulation models, 
because many of the mechanistic details of coagulation were encoded as logical rules.
We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
We then validated the model ensemble using thrombin data sets that were not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems.

@@

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics, and encode mechanism.
However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, it is not clear whether we actually need precise estimates for all model parameters. 
Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well defined. 
Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
without requiring mechanism \citep{Morris:2010aa}. 
Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
However, modeling complex dynamics with logical models is challenging. 
Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
while missing or incomplete mechanistic knowledge can be approximated using a logical approach.

In this study, we developed a hybrid approach which combined ODE modeling with logical rules to model a well studied biochemical network, 
the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by TF/FVIIa \cite{Butenas:2002aa}. 
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
leading to the formation of the tenase and prothrombinase complexes on activated platelets.
These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
neutralizes several of the proteases generated during coagulation, including thrombin.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass-action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

%However, components could also correspond to non-modeled influences, categorial variables or other abstract quantities.
%In the current proof-of-principle model, 
%APC can inhibit amplification, and upstream of amplification thus APC activity will directly slow the rate of thrombin formation from the initiation trigger, and amplification itself.
%On the other hand, ATIII inhibits thrombin directly.

@@

The reduced order coagulation model captured the role of initial prothrombin abundance, and the decay of the thrombin signal following from ATIII activity (Fig. \ref{fig-diff_fII_NO_APC}).
However, we systematically under-predicted the thrombin peak and the strength of ATIII inhibition in this training data set.
On the other hand, with fixed thrombin parameters, we captured peak thrombin values and the decay of the thrombin signal (at least for the 150\% fII case) in the presence
of both ATIII and the protein C pathway (Fig. \ref{fig-diff_fII_APC}). Lastly, we were unable to capture global differences in initiation time \textit{across} separate data 
sets with a single ensemble of model parameters. These differences likely resulted from normal experimental variability.
For example, different thrombin generation experiments within the 
training data (at the same physiological factor levels) had significantly different initiation times (data not shown). 
However, the inability to globally capture initiation time also highlighted a potential shortcoming of the initiation module within the model. 
To capture the variability in initiation time \textit{across} training data sets, we included a constant time-delay parameter ($T_{D}$) for each data group.
The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. 
Introduction of the delay parameter allowed the model to simulate multiple training data sets using a single ensemble of model parameters.
Taken together, the model identification results suggested 
that our hybrid approach could reproduce a panel of thrombin generation data sets in the neighborhood of physiological factor and inhibitors concentrations.
However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.

\subsection*{Validation of the reduced order coagulation model.}
We tested the predictive power of the reduced order coagulation model with validation data sets not used during model training. 
Two validation data sets were used, thrombin generation for various prothrombin and ATIII concentrations with the protein C pathway, 
and thrombin generation in normal versus hemophilic plasma in the presence of the protein C pathway. 
Lastly, we compared the qualitative output of the model to rFVIIa addition in the presence of hemophilia.
The hemophilia case was an especially difficult test as it was taken from a different study which used a plasma-based
\textit{in~vitro} assay involving platelets instead of phospholipid vesicles (PCPS). 
All kinetic and control parameters were fixed for the validation simulations. 
The only globally adjustable parameter $T_{D}$, was fixed within each validation data set but allowed to vary between data sets.
The reduced order model predicted the thrombin generation profile for ratios of prothrombin and ATIII in the absence of the protein C pathway (Fig. \ref{fig-fII_ATIII}).
Simulations near the physiological range (fII,ATIII) = (100\%, 100\%) or (125\%,75\%) tracked the measured thrombin values (Fig. \ref{fig-fII_ATIII}B and C).
On the other hand, predictions for factors levels outside of the physiological range (fII,ATIII) = (50\%, 150\%) or (150\%, 50\%), while qualitatively consistent with measured thrombin values, did show significant 
deviation from the measurements (Fig. \ref{fig-fII_ATIII}A and D).
Likewise, simulations of thrombin generation in normal versus hemophilia (missing both fVIII and fIX) were consistent with measured thrombin values (Fig. \ref{fig-hemophilia}).
We modeled the dependence of thrombin amplification on factor levels using a product rule ($\mathcal{Z} = fV\times fX\times fVIII\times fIX$), 
which was then was integrated using a $\min$ integration rule into the control variable governing amplification. 
Thus, in the absence of fVIII or fIX, the amplification control variable evaluated to zero, and the only thrombin produced was from initiation (Fig. \ref{fig-hemophilia}B).
However, the decay of the thrombin signal was underpredicted in the normal case (Fig. \ref{fig-hemophilia}A), while the activated thrombin level was overpredicted 
in hemophilia simulations, although thrombin generation was far less than normal (Fig. \ref{fig-hemophilia}B). 
Taken together, the reduced order model performed well in the physiological range of factors, even with
unmodeled components such as platelet activation in the hemophilia data set.

The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa versus rFVIIa, as we have only a single lumped initiation source (trigger).
Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin produced was from initiation (Fig. \ref{fig-rFVII}A, 1$\times$~trigger).
However, as we increased the trigger strength, the thrombin profile began to approximate normal coagulation, showing a pronounced peak albeit with a slower peak time (Fig. \ref{fig-rFVII}B).
Further increases in trigger strength resulted in decreased thrombin peak time and increased maximum thrombin values (Fig. \ref{fig-rFVII}A, 50$\times$~trigger). 
Thus, for large trigger values (200$\times$trigger) the hemophilic thrombin profile approximated normal coagulation, where peak thrombin was achieved shortly 
after administration and 95\% of the thrombin was gone by 20 min after initiation. 
We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig. \ref{fig-flux-analysis}).
We calculated the flux through four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
For normal coagulation, amplification and thrombin inhibition were the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
After 10 min, the dominate reaction had shifted to thrombin inhibition (both ATIII and APC action). 
In hemophilia (missing both fVIII and fIX), the amplification reaction did not occur, and thrombin was produced only by initiation (Fig. \ref{fig-flux-analysis}, center).
Initiation was quickly inhibited by APC, and the thrombin level stabilized (eventually decaying at longer times because of ATIII activity). 
Lastly, when 50$\times$trigger was used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
initiation mechanisms dominated for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
Similar to hemophilia alone, no amplification occurred in the 50$\times$trigger+hemophilia case, 
and the rate of thrombin generation was extinguished by the combined action of ATIII and APC. 
Taken together, the hybrid modeling approach captured the transition between the modes of thrombin generation, as well as the role that inhibitors play 
in attenuating the thrombin generation rate. Thus, the transfer function approach encoded the inhibitory logic of this cascade in the absence of specific mechanism.

@@

\subsection*{Global sensitivity analysis of the reduced order coagulation model.}
We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
We calculated the sensitivity of the time to maximum thrombin (peak time) and the thrombin exposure (area under the thrombin curve) 
for different levels of prothrombin, and protein C (Fig. \ref{fig-sensitivity}). 
Globally, 41\% of the parameters shifted in importance between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases for the peak thrombin time (Fig. \ref{fig-sensitivity}A). 
The majority of these shifts involved the interaction between increased prothrombin and the protein C pathway, while only
5\% were directly associated with increased prothrombin alone. The rate constant for thrombin amplification was the most important parameter controlling the
peak thrombin time. While this parameter was differentially important for different prothrombin levels, and in the presence or absence of 
the activated protein C pathway, it's was consistently the most sensitive parameter in the model. The saturation constant governing thrombin amplification was
the second most important parameter, followed by the initiation control gain parameter. Other important parameters influencing the thrombin peak time included the control gain for
activated protein C formation, and the rate constant controlling ATIII inhibition of thrombin activity. 
On the other hand, only 27\% of the model parameters were differentially sensitive between the (fII,PC) = (50\%, 0\%) and (150\%,100\%) cases 
for thrombin exposure (Fig. \ref{fig-sensitivity}B). Of these parameters, all of the shifts were associated with the interplay between thrombin formation and the protein C pathway. 
The rate constant controlling ATIII inhibition was the most important parameter controlling the thrombin exposure.
While this parameter was less important in the presence of protein C for 150\% prothrombin levels, it was significantly above all other parameters. 
Similar to the peak time, for 150\% prothrombin, the control gain for activated protein C formation was differentially important along with the rate constant controlling
amplification. However, the amplification parameter was much less important for thrombin exposure versus peak time.
